Modelling state-switching functional data with hidden Markov models

### 2020-10-30 Report/CJS_Sample.tex
commit before adding t_tilde

% Abstract, keywords, and classification codes
\startabstract{%
\keywords{
\KWDtitle{Key words and phrases}
Association parameters\sep clustered data\sep mean parameters\sep missing data\sep pairwise
likelihood\sep repeated measurements.
% MSC 2010 subject classification codes
\KWDtitle{MSC 2010}Primary 62???\sep secondary 62???}
\begin{abstract}
\abstractsection{}
\abstractsection{Abstract}
Insert your abstract here; it should
typically be up to ten lines long. Avoid symbols as much as possible. Formulas
are strongly discouraged, and citations should be avoided. 
The title and the abstract should be concise and descriptive;
list the key words in alphabetical order.  The MSC 2010 subject classification codes can be found here:
http://www.ams.org/mathscinet/msc/pdfs/classifications2010.pdf.
\abscopyright
\fabstractsection{}
\fabstractsection{R\'{e}sum\'{e}}
Ins\'{e}rer votre r\'{e}sum\'{e}
ici. We will supply a French abstract for those authors who
can't prepare it themselves.\Frenchabscopyright
\end{abstract}}
\makechaptertitle

>>>>

% Abstract, keywords, and classification codes
\startabstract{%
\keywords{
\KWDtitle{Key words and phrases}
Hidden Markov Models\sep biologging\sep multi-state model\sep missing data\sep pairwise
likelihood\sep repeated measurements.
% MSC 2010 subject classification codes
\KWDtitle{MSC 2010}Primary 62M05 \sep secondary 62P12}
\begin{abstract}
\abstractsection{}
\abstractsection{Abstract}
Insert your abstract here; it should
typically be up to ten lines long. Avoid symbols as much as possible. Formulas
are strongly discouraged, and citations should be avoided. 
The title and the abstract should be concise and descriptive;
list the key words in alphabetical order.  The MSC 2010 subject classification codes can be found here:
http://www.ams.org/mathscinet/msc/pdfs/classifications2010.pdf.
\abscopyright
\fabstractsection{}
\fabstractsection{R\'{e}sum\'{e}}
Ins\'{e}rer votre r\'{e}sum\'{e}
ici. We will supply a French abstract for those authors who
can't prepare it themselves.\Frenchabscopyright
\end{abstract}}
\makechaptertitle

### 2020-10-30 Report/Sections/appendix.tex
commit before adding t_tilde

\citet{Michelot:2019} model the movement of an animal as the solution to the following stochastic differential equation:
%
\begin{equation}
    \label{eqn:SDE}
    dY_t = \beta^{(X_t)}(\gamma^{(X_t)} - Y_t)dt + \omega^{(X_t)} dW_t,
\end{equation}
%
where $X_t$ is some stochastic process which defines the hidden behaviour of the animal at time $t$, $\beta^{(X_t)}$ relates to rate at which the process returns to its mean value, $\gamma^{(X_t)}$ is the long-term mean value of the process in state $X_t$, $\omega^{(X_t)}$ is related to short-term variance, and $W$ is a Wiener process. They refer to this as a state-switching Ornstein-Uhlenbeck process. Unlike an HMM, $t \in \mathbb{R}$ indexes a continuous time process and is therefore not necessarily an integer in this case. If the behavioural state $X_t$ is known and does not change between observations $\Big($i.e $X_s = X_t$ for all $s \in [t,t+\Delta t)\Big)$, the solution to Equation (\ref{eqn:SDE}) is known to be the following \citep{Michelot:2019}:
%
\begin{align}
    Y_{t+\Delta t} | X_{t} \sim \mathcal{N}\left((1-e^{-\beta^{(X_t)}\Delta t})\gamma^{(X_t)} + e^{-\beta^{(X_t)}\Delta t} Y_t,\quad \frac{\omega^{(X_t)^2}}{2\beta^{(X_t)}} (1-e^{-2\beta^{(X_t)}\Delta t})\right)
    \label{eqn:OU_sol}
\end{align}
%
where $\Delta t$ is the time difference between any two observations $Y_t$ and $Y_{t+\Delta t}$. In fact, under certain conditions, the output of a state-switching OU process is identical to that of a CarHMM, as described by Theorem 1 below.

    \item Observations of the SDE from Equation (\ref{eqn:OU_sol}) are taken at times $\left(0, \Delta t, \ldots, (T-1)\Delta t\right)$,
\end{enumerate}
then the observations $Y$ are equivalent to the output of a conditionally auto-regressive hidden Markov model with normal emission distributions and parameters $\theta = (\theta^{(1)}, \ldots, \theta^{(N)}); \enspace \theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where:

@@

It is straightforward to simulate realizations of the coarse-scale HMM ($X$ and $Y$) given the parameters $\Gamma$ and $\theta$.

To simulate $X^*_t$ for each dive $t$, note that we simulate accelerometer readings taken at 50 hertz and have two-second window lengths. Therefore, each window contains 100 accelerometer readings and $X^*_t = \left\{X^*_{t,1}, X^*_{t,101}, \ldots, X^*_{t,100T^*_t-99}\right\}$. In addition, the length of each fine-scale Markov chain is $T^*_t = \lfloor Y_t/2 \rfloor$. Then, given the dive types $X_t$, dive durations $Y_t$, and the fine-scale probability transition matrices $\Gamma^{*(1)}$ and $\Gamma^{*(2)}$, it is straightforward to simulate $X^*_t$ for each dive $t = 1,2,\ldots,100$. Recall that $\Gamma^{*(i)}_{jk} = Pr(X^*_{t,t^*+100} = k | X^*_{t,t^*} = j, X_t = i)$ for all $t$ and $t^* = 1, 101, \ldots$. It is much more difficult to simulate the raw acceleration data $Y^*$, and we focus on this task for the remainder of this section.

@@

>>>>

@@

It is straightforward to simulate realizations of the coarse-scale HMM ($X$ and $Y$) given the parameters $\Gamma$ and $\theta$. 
To simulate $X^*_t$ for each dive $t$, note that we simulate accelerometer readings taken at 50 hertz and have two-second window lengths. Therefore, each window contains 100 accelerometer readings and $X^*_t = \left\{X^*_{t,1}, X^*_{t,101}, \ldots, X^*_{t,100T^*_t-99}\right\}$. In addition, the length of each fine-scale Markov chain is $T^*_t = \lfloor Y_t/2 \rfloor$. Then, given the dive types $X_t$, dive durations $Y_t$, and the fine-scale probability transition matrices $\Gamma^{*(1)}$ and $\Gamma^{*(2)}$, it is straightforward to simulate $X^*_t$ for each dive $t = 1,2,\ldots,100$. Recall that $\Gamma^{*(i)}_{jk} = Pr(X^*_{t,t^*+100} = k | X^*_{t,t^*} = j, X_t = i)$ for all $t$ and $t^* = 1, 101, \ldots$. It is much more difficult to simulate the raw acceleration data $Y^*$, and we focus on this task for the remainder of this section.

@@

\iffalse

\citet{Michelot:2019} model the movement of an animal as the solution to the following stochastic differential equation:
%
\begin{equation}
    \label{eqn:SDE}
    dY_t = \beta^{(X_t)}(\gamma^{(X_t)} - Y_t)dt + \omega^{(X_t)} dW_t,
\end{equation}
%
where $X_t$ is some stochastic process which defines the hidden behaviour of the animal at time $t$, $\beta^{(X_t)}$ relates to rate at which the process returns to its mean value, $\gamma^{(X_t)}$ is the long-term mean value of the process in state $X_t$, $\omega^{(X_t)}$ is related to short-term variance, and $W$ is a Wiener process. They refer to this as a state-switching Ornstein-Uhlenbeck process. Unlike an HMM, $t \in \mathbb{R}$ indexes a continuous time process and is therefore not necessarily an integer in this case. If the behavioural state $X_t$ is known and does not change between observations $\Big(X_s = X_t$ for all $s \in [t,t+\Delta t)\Big)$, the solution to Equation (\ref{eqn:SDE}) is known to be the following \citep{Michelot:2019}:
%
\begin{align}
    Y_{t+\Delta t} | X_{t} \sim \mathcal{N}\left((1-e^{-\beta^{(X_t)}\Delta t})\gamma^{(X_t)} + e^{-\beta^{(X_t)}\Delta t} Y_t,\quad \frac{\omega^{(X_t)^2}}{2\beta^{(X_t)}} (1-e^{-2\beta^{(X_t)}\Delta t})\right),
    \label{eqn:OU_sol}
\end{align}
%
where $\Delta t$ is the time difference between any two observations $Y_t$ and $Y_{t+\Delta t}$. In fact, under certain conditions, the output of a state-switching OU process is identical to that of a CarHMM, as described by Theorem 1 below.

    \item Observations of the SDE from Equation (\ref{eqn:OU_sol}) are taken at times $\left(0, \Delta t, \ldots, (T-1)\Delta t\right)$,
\end{enumerate}
then the observations $Y$ are equivalent to the output of a conditionally auto-regressive hidden Markov model with normal emission distributions and parameters $\theta = \{\theta^{(1)}, \ldots, \theta^{(N)}\}; \enspace \theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$, where:

\fi

### 2020-10-30 Report/Sections/models.tex
commit before adding t_tilde

The CarHMM, described by \citet{Lawler:2019}, is a generalization of an HMM which explicitly models auto-correlation in the observation sequence beyond the correlation induced by the hidden state process. Like a traditional HMM, a CarHMM is made up of a Markov chain of unobserved states $X_1,\ldots, X_T$ that can take on values $1, \ldots, N$, with transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$. Unlike a traditional HMM, the CarHMM assumes that the distribution of $Y_t$ conditioned on $X_1,\ldots, X_T$ and $Y_1,\ldots, Y_{t-1}$ depends on both $X_t$ and $Y_{t-1}$ rather than only $X_t$. 
The first emission $Y_1$ is assumed to be fixed and known as an initial value which does not depend upon $X_1$.
%
We denote the conditional density of $Y_t$ given $Y_{t-1} = y_{t-1}$ and $X_t=i$ as $f^{(i)}( \cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}( \cdot | y_{t-1})$.
For example, if $Y_t$ is a scalar, one could assume that its conditional distribution is Normal with parameters $\theta^{(i)} = \{\mu^{(i)},\sigma^{(i)},\phi^{(i)}\}$ where:
%
\begin{align}
\label{eqn:carhmm}
\begin{split}
\mathbb{E}(Y_t|Y_{t-1} = y_{t-1},X_t=i) &= \phi^{(i)} ~ y_{t-1} ~+ ~(1-\phi^{(i)})  ~\mu^{(i)}, \text{ and } \\
\mathbb{V}(Y_t| Y_{t-1} = y_{t-1}, X_t = i) &= (\sigma^{(i)})^2.
\end{split}
\end{align}
%
If $Y_t$ is a vector, Equation (\ref{eqn:carhmm}) above can model either all dimensions of $Y_t$ simultaneously or just a subset of the dimensions, depending upon considerations such as subject mater expertise or expository data analysis. Figure \ref{fig:models}b shows a graphical representation of the dependence structure of a CarHMM.

The likelihood for the CarHMM can be easily calculated using the forward algorithm. As previously, let $y$ be the vector of observed emissions. Then:
\begin{equation*}
    \calL_{\text{CarHMM}}(\theta,\Gamma;y) = \delta \prod_{t=2}^T \Gamma P(y_t|y_{t-1};\theta) \mathbf{1}_N,
    \label{eqn:CarHMM_likelihood}
\end{equation*}
where
%
$P(y_t|y_{t-1};\theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to $f^{(i)}(y_t|y_{t-1}; \theta^{(i)})$.

@@

An HHMM accounts for different scales of correlation by modelling both the coarse-scale process and fine-scale process with distinct HMMs \citep{Barajas:2017,Adam:2019}. The coarse-scale model is an HMM as defined previously, where $X_1, \ldots, X_T$ make up an unobserved Markov chain with $N$ possible states and $Y_1,\ldots, Y_T$ are the corresponding observations.   
%
In the hierarchical setting, each state $X_t$ also emits another sequence of fine-scale unobserved states, $X_t^* \equiv \{X_{t,1}^*,\ldots, X_{t,T_t^*}\}$, which in turn emits a sequence of fine-scale observations $Y_t^* \equiv \{Y_{t,1}^*,\ldots, Y_{t,T_t^*}\}$. For each $t$, the fine-scale process $\{X_t^*, Y_t^*\}$ then follows another HMM whose parameters depend on the value of $X_t$. Specifically, if $X_t=i$, then the components of $X_t^*$ are a Markov chain with an $N^*_t \times N^*_t$ transition probability matrix $\Gamma^{*(i)}$ and initial distribution $\delta^{*(i)}$, which we assume is equal to the stationary distribution of the chain. For simplicity, we take $N_t^* \equiv N^*$ although this is not necessary in general. The distribution of $Y_{t,t^*}$ given $X_{t,t^*}=i^*$ and $X_t=i$ is governed by the parameter $\theta^{*(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot; \theta^{*(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot)$. We denote the fine-scale emission parameter vector corresponding to $X_t=i$ as $\theta^{*(i)}=\left(\theta^{*(i,1)}, \ldots, \theta^{*(i,N^*)}\right)$. Given the coarse-scale hidden states, the $T$ fine-scale HMMs $\{X_1^*, Y_1^*\}, \ldots, \{X_T^*, Y_T^*\}$, are all assumed to be independent of one another.

Depending upon the process being modelled, it is possible to force certain parameters to be shared to reduce the complexity of the HHMM. For example, in the killer whale case study (see Section \ref{sec:data}), we force the fine-scale emission parameters to be shared across coarse-scale hidden states $\left( \text{i.e. } \theta^{*(1,i^*)} = \cdots = \theta^{*(N,i^*)} \text{ for all } i^* = 1, \ldots, N^* \right)$. Figure \ref{fig:models}c represents the dependence structure for an HHMM.

@@

In many applications where data are collected at high frequencies, intricate dependency structures may arise within the fine-scale process which cannot be adequately modelled with any of the HMM-based models mentioned thus far. For example, the accelerometer data in Figure \ref{fig:data} sporadically displays clear periodic behaviour. To handle these additional fine-scale structures, we recommend replacing the fine-scale observations $Y_t^*$ with relevant statistics that summarize this non-Markovian behaviour. To maintain the temporal structure of the fine-scale process, summary statistics can be calculated from a moving window over $Y_t^*$. Subject-area experts are often required to determine the optimal window size and the specific summary statistics used. For our case study, we use the discrete Fourier transform (DFT) of a moving forward window of width $h$ across $Y^*_t$:
%
\begin{align*}
    DFT\{Y^*_{t,t^*}\}(k) = \sum_{n=0}^{h-1} Y^*_{t,t^*+n}\exp\left(-\frac{i 2\pi}{h} kn \right),
\end{align*}
%
for $k = 0, 1, \ldots, h-1$. If $Y^*_{t,t^*}$ is high-dimensional, then the DFT is taken component-wise. Next, we calculate $\Z_{t,t^*} \equiv \{\Zone_{t,t^*},\Ztwo_{t,t^*}\}$ for each moving window, where
%
\begin{equation}
    \label{eqn:z}
    \Zone_{t,t^*} \equiv \frac{1}{h}\sum_{n=0}^{h-1}Y^*_{t,t^*+n} \quad \text{and} \quad \Ztwo_{t,t^*} \equiv \sum_{k=1}^{\tilde{\omega}}\bigg|\bigg|DFT\{Y^*_{t,t^*}\}(k)\bigg|\bigg|^2.
\end{equation}
%
In words, $\Zone_{t,t^*}$ is the average value over the window and $\Ztwo_{t,t^*}$ is the squared two-norm of the component of the window $\left(Y^*_{t,t^*}, \ldots, Y^*_{t,t^*+h-1}\right)$ that can be attributed to frequencies between one and $\tilde{\omega}$ periods per window (by the Plancherel theorem). More intuitively, $\Ztwo_{t,t^*}$ corresponds to the ``wiggliness" of the acceleration data for dive $t$ within the window starting at time stamp $t^*$. The maximum frequency $\tilde{\omega}$ is a problem-specific tuning parameter which should be selected with the help of a subject-matter expert. Other DFT-based choices for window summary statistics are possible, including the dominant frequency and amplitude.

To avoid introducing auto-correlation into the transformed observation sequence, we recommend down-sampling $\Z_t$ such that $\Z_t = \left\{\Z_{t,1},\Z_{t,h+1},\Z_{t,2h+1},\ldots\right\}$ and $X^*_t = \left\{X^*_{t,1},X^*_{t,h+1},X^*_{t,2h+1},\ldots\right\}$. When down-sampling, we redefine the fine-scale probability transition matrices such that $\Gamma^{*(i)}_{jk} = \Pr(X^*_{t,t^*+h} = k | X^*_{t,t^*} = j, X_t = i)$ for all $t$ and $t^* = 1,h+1,2h+1,\ldots$. %In addition, we assume that the hidden state within a window does not change, (i.e. $X^*_{t,t^*+s} = X^*_{t,t^*}$ for all $t$, $t^* = 1,h+1,2h+1,\ldots$ and $s = 0,1,\ldots,h-1$. 
This down-sampling both avoids the artificial residual correlation that overlapping windows would introduce and reduces the dimension of the fine-scale process by a factor of $h$, allowing for faster model fitting. By summarizing the data using a relevant transformation, the method overall also reduces correlation in the raw observations due to the complex dependence structure. Figure \ref{fig:models}d represents the process of transforming $Y^*_{t}$ into $\Z_{t}$ within an HMM, including down-sampling.

@@

On the fine scale, $Y^*_t$ can be modelled using a large variety of parametric models which admit easy-to-compute likelihoods or penalized likelihoods. In this case, the fine-scale HMM likelihood term $\calL_{\text{HMM}}$ in Equation (\ref{eqn:HHMM_likelihood}) is replaced by the likelihood of a general fine-scale model, $\calL_{\text{fine}}(\theta^{*(i)};y^*_t)$:
\[
\calL_{\text{coarse}}(\theta,\theta^*,\Gamma;y,y^*) = \delta P(y_1,y^*_1;\theta,\theta^*) \prod_{t=2}^T \Gamma P(y_t,y^*_t;\theta,\theta^*) \mathbf{1}_N
\]
where $P(y_t,y^*_t;\theta,\theta^*) $ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t;\theta^{(i)}\right)\calL_{\text{fine}}\left(\theta^{*(i)};y^*_t\right)$. 
%
Possible candidates for the fine-scale model include any of the models described in the previous subsections (HMM, CarHMM, HHMM, and HHMM-DFT) in addition to many others not described here. 
For example, a continuous-time HMM (CTHMM) may be appropriate if observations are not equi-spaced in time \citep{Liu:2015}. as an example, \citet{Xu:2018} explain biologging accelerometer data by incorporating a CTHMM into a hierarchical model similar to ours. However, they take a longitudinal approach whereas we use a coarse-scale HMM in our case study.
Further, \citet{Bebbington:2007} and \citet{Borchers:2013} investigate data sets with count onsets as observations, so they use variations on a Poisson process as their fine-scale model. If the fine-scale model is a simple Poisson process, then the hierarchical structure described here reduces to a Markov-modulated Poisson process \citep{Fischer:1993}.
As a final example, \citep{Langrock:2018} use B-splines to model the emission distribution of an HMM as well as incorporate covariates. This non-parametric approach admits a penalized likelihood which can easily be incorporated into the hierarchical structure here.

These examples are a few of many fine-scale models that can act as initial building blocks in a practitioner's toolbox to construct increasingly complex hierarchical models. As an example, consider a \textbf{CarHHMM}, in which the coarse-scale model is a traditional HMM and the fine-scale model is a CarHMM. Alternatively, one may construct an \textbf{HHHMM}, where three scales of observations are recorded ($Y$, $Y^*$, and $Y^{**}$) and traditional HMMs are used to model all three levels. It is clear that are a myriad of possible models can be built using this framework. However, these models quickly become complicated and should be constructed with care to achieve an adequate fit of the data while avoiding over-fitting and high computational costs.

>>>>

The CarHMM, described by \citet{Lawler:2019}, is a generalization of an HMM which explicitly models auto-correlation in the observation sequence beyond the correlation induced by the hidden state process. Like a traditional HMM, a CarHMM is made up of a Markov chain of unobserved states $X_1,\ldots, X_T$ that can take on values $1, \ldots, N$, with transition probability matrix $\Gamma$ and initial distribution $\delta$ equal to the stationary distribution of $\Gamma$. Unlike a traditional HMM, the CarHMM assumes that the distribution of $Y_t$ conditioned on $X_1,\ldots, X_T$ and $Y_1,\ldots, Y_{t-1}$ depends on both $X_t$ and $Y_{t-1}$ rather than only $X_t$. The first emission $Y_1$ is assumed to be fixed and known as an initial value which does not depend upon $X_1$. We denote the conditional density of $Y_t$ given $Y_{t-1} = y_{t-1}$ and $X_t=i$ as $f^{(i)}(\cdot | y_{t-1}; \theta^{(i)})$ or simply $f^{(i)}(\cdot | y_{t-1})$. This model is highly general, as $f^{(i)}(\cdot | y_{t-1})$ can be any valid density that depends upon the parameters $\theta^{(i)}$ and the previous observation $Y_{t-1}$. 
As a concrete example, if $Y_t$ is a $K$-dimensional vector with components $Y_t(1),\ldots,Y_t(K)$, consider the component-wise sequences $Y_1(k), \ldots, Y_T(k)$ for $k = 1,\ldots K$. One may assume that these sequences are independent given $X_1,\ldots,X_T$ with $Y_t(k)|X_t = i$ and that $Y_t(k)$ given $X_t = i$ is Normally distributed with parameters $\theta^{(i)}_k = \{\mu^{(i)}_k,\sigma^{(i)}_k,\phi^{(i)}_k\}$, where:
%
\begin{align}
\label{eqn:carhmm}
\begin{split}
\mathbb{E}(Y_{t}(k)|Y_{t-1}(k) = y_{t-1}(k),X_t=i) &= \phi^{(i)}_k ~ y_{t-1}(k) ~+ ~(1-\phi^{(i)}_k)  ~\mu^{(i)}_k, \\
\mathbb{V}(Y_t(k)| Y_{t-1}(k) = y_{t-1}(k), X_t = i) &= (\sigma^{(i)}_k)^2.
\end{split}
\end{align}

A CarHMM which follows Equation (\ref{eqn:carhmm}) can be viewed as a discrete time version of the state-switching Ornstein-Uhlenbeck process of \citet{Michelot:2019} if observation error is negligible. This follows in the same way that an AR(1) process is the discrete-time version of a traditional Ornstein-Uhlenbeck process. Figure \ref{fig:models}b shows a graphical representation of the dependence structure of a general CarHMM. The likelihood corresponding to a general CarHMM can be easily calculated using the forward algorithm. As previously, let $y$ be the vector of observed emissions. Then:
\begin{equation*}
    \calL_{\text{CarHMM}}(\theta,\Gamma;y) = \delta \prod_{t=2}^T \Gamma P(y_t|y_{t-1};\theta) \mathbf{1}_N,
    \label{eqn:CarHMM_likelihood}
\end{equation*}
where
%
$P(y_t|y_{t-1};\theta)$ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry equal to $f^{(i)}(y_t|y_{t-1}; \theta^{(i)})$.

@@

An HHMM accounts for different scales of correlation by modelling both the coarse-scale process and fine-scale process with either HMMs \citep{Barajas:2017,Adam:2019} or CarHMMs. The coarse-scale model is as defined previously, where $X_1, \ldots, X_T$ make up an unobserved Markov chain with $N$ possible states and $Y_1,\ldots, Y_T$ are the corresponding observations with state-dependent parameters $\theta^{(i)}, i = 1,\ldots,N$.   
%
In the hierarchical setting, each state $X_t$ also emits another sequence of fine-scale unobserved states, $X_t^* \equiv \{X_{t,1}^*,\ldots, X_{t,T_t^*}\}$, which in turn emits a sequence of fine-scale observations $Y_t^* \equiv \{Y_{t,1}^*,\ldots, Y_{t,T_t^*}\}$. For each $t$, the fine-scale process $\{X_t^*, Y_t^*\}$ then follows another HMM (or CarHMM) whose parameters depend on the value of $X_t$. If $X_t = i$, then the components of $X_t^*$ are a Markov chain with an $N^{*(i)} \times N^{*(i)}$ transition probability matrix $\Gamma^{*(i)}$ and initial distribution $\delta^{*(i)}$, which we assume is equal to the stationary distribution of the chain. The distribution of $Y_{t,t^*}$ given $Y_{t,t^*-1} = y_{t,t^*-1}$, $X_{t,t^*}=i^*$, and $X_t=i$ is governed by the parameter $\theta^{*(i,i^*)}$ and has density or probability mass function denoted $f^{*(i,i^*)}\left(\cdot|y^*_{t,t^*-1}; \theta^{*(i,i^*)}\right)$ or simply $f^{*(i,i^*)}(\cdot|y^*_{t,t^*-1})$. We denote the fine-scale emission parameter vector corresponding to $X_t=i$ as $\theta^{*(i)}=\left(\theta^{*(i,1)}, \ldots, \theta^{*(i,N^{*(i)})}\right)$. In summary:
%
% this is new
%
\begin{gather*}
    \{Y, X\} \text{ follows (Car)HMM with } \Gamma \in \bbR^{N \times N}, \\
    %
    \{Y_t   | Y_{t-1} = y_{t-1}, X_t = i\} \sim f^{(i)}(\cdot|y_{t-1};\theta^{(i)}), \\
    %
    \{Y^*_t,X^*_t | X_t = i\} \text{ follows (Car)HMM with } \Gamma^{*(i)} \in \bbR^{N^{*(i)} \times N^{*(i)}}, \\
    %
    \{Y^*_{t,t^*} | Y^*_{t,t^*-1} = y^*_{t,t^*-1}, X^*_{t,t^*} = i^*, X_t = i\} \sim f^{*(i,i^*)}(\cdot|y^{*}_{t,t^*-1};\theta^{(i,i^*)}).
\end{gather*}
Given the coarse-scale hidden states $X$, the $T+1$ sets $\{X_1^*, Y_1^*\}, \ldots, \{X_T^*, Y_T^*\}$ and $\{Y_1,\ldots,Y_T\}$ are assumed to be independent of one another. Figure \ref{fig:models}c graphically displays the dependence structure for an HHMM.
%
% end new
%

Forcing certain parameters to be shared can reduce the complexity of the HHMM. For example, in the killer whale case study (see Section \ref{sec:data}), we take $N^{*(i)} = N^*$ for all $i$. We also share the fine-scale emission parameters across coarse-scale hidden states $\left( \text{i.e., } \theta^{*(1,i^*)} = \cdots = \theta^{*(N,i^*)} = \theta^{*(\cdot,i^*)} \text{ for all } i^* = 1, \ldots, N^* \right)$.

@@

In many applications where data are collected at high frequencies, intricate dependency structures may arise within the fine-scale process which cannot be adequately modelled with any of the HMM-based models mentioned thus far. For example, the accelerometer data in Figure \ref{fig:data} sporadically displays clear periodic behaviour. To handle these additional fine-scale structures, we recommend replacing $Y_t^* = \{Y^*_{t,1},\ldots,Y^*_{t,T^*_t}\}$ with relevant statistics that summarize this non-Markovian behaviour. To maintain the temporal structure of the fine-scale process, summary statistics can be calculated from a moving window over the elements $Y_t^*$. Subject-area experts are often required to determine the optimal window size and the specific summary statistics used. For our case study, we use the discrete Fourier transform (DFT) of a moving forward window of width $h$ across $Y^*_t$. For each $t^* = 1,2,\ldots,T^*_t-h+1$:
%
\begin{align*}
    DFT\{Y^*_{t,t^*},\ldots, Y^*_{t,t^*+h-1}\}(k) = \sum_{n=0}^{h-1} Y^*_{t,t^*+n}\exp\left(-\frac{i 2\pi}{h} kn \right) \qquad 
\end{align*}
%
for $k = 0, 1, \ldots, h-1$. If $Y^*_{t,t^*}$ is a vector then the DFT is taken component-wise. Next, we calculate $\Z_{t,t^*} \equiv \{\Zone_{t,t^*},\Ztwo_{t,t^*}\}$ for each moving window, where
%
\begin{equation}
    \label{eqn:z}
    \Zone_{t,t^*} \equiv \frac{1}{h}\sum_{n=0}^{h-1}Y^*_{t,t^*+n}, \quad \Ztwo_{t,t^*} \equiv \sum_{k=1}^{\tilde{\omega}}\Big|\Big|DFT\{Y^*_{t,t^*},\ldots, Y^*_{t,t^*+h-1}\}(k)\Big|\Big|^2.
\end{equation}
%
In words, $\Zone_{t,t^*}$ is the average value over the window and $\Ztwo_{t,t^*}$ is the squared two-norm of the component of the window $\big\{Y^*_{t,t^*}, \ldots, Y^*_{t,t^*+h-1}\big\}$ that can be attributed to frequencies between one and $\tilde{\omega} \leq h-1$ periods per window (by the Plancherel theorem). More intuitively, $\Ztwo_{t,t^*}$ corresponds to the ``wiggliness" of the acceleration data for dive $t$ within the window starting at time stamp $t^*$. The maximum frequency $\tilde{\omega}$ is a problem-specific tuning parameter which should be selected with the help of a subject-matter expert. Other DFT-based choices for window summary statistics are possible, including the dominant frequency and amplitude.

To avoid introducing auto-correlation into the transformed observation sequence, we recommend down-sampling $\Z_t$ such that $\Z_t = \big\{\Z_{t,1},\Z_{t,h+1},\Z_{t,2h+1},\ldots\big\}$ and $X^*_t = \big\{X^*_{t,1},X^*_{t,h+1},X^*_{t,2h+1},\ldots\big\}$. When down-sampling, we redefine the fine-scale probability transition matrices such that $\Gamma^{*(i)}_{jk} = \Pr(X^*_{t,t^*+h} = k | X^*_{t,t^*} = j, X_t = i)$ for all $t$ and $t^* = 1,h+1,2h+1,\ldots$. If the fine-scale model is a CarHMM, we also redefine the probability density function $f$ such that $f^{*(i,i^*)}(\cdot|\z_{t,t^*-100}; \theta^{*(i,i^*)})$ describes the density of $\Z_{t,t^*}$ given $X_t = i$, $X^*_{t,t^*} = i^*$, and $\Z_{t,t^*-100} = \z_{t,t^*-100}$.
This down-sampling both avoids the artificial residual correlation that overlapping windows would introduce and reduces the dimension of the fine-scale process by a factor of $h$, allowing for faster model fitting. By summarizing the data using a relevant transformation, the method overall also reduces correlation in the raw observations due to the complex dependence structure. Figure \ref{fig:models}d represents the process of transforming $Y^*_{t}$ into $\Z_{t}$ within an HMM, including down-sampling.

@@

On the fine scale, $Y^*_t$ can be modelled using a large variety of parametric models which admit easy-to-compute likelihoods or penalized likelihoods. In this case, the fine-scale HMM likelihood term $\calL_{\text{HMM}}$ in Equation (\ref{eqn:HHMM_likelihood}) is replaced by the likelihood of a general fine-scale model, $\calL_{\text{fine}}(\theta^{*(i)};y^*_t)$:
\[
\calL_{\text{coarse}}(\theta,\theta^*,\Gamma;y,y^*) = \delta P(y_1,y^*_1;\theta,\theta^*) \prod_{t=2}^T \Gamma P(y_t,y^*_t;\theta,\theta^*) \mathbf{1}_N
\]
where $P(y_t,y^*_t;\theta,\theta^*) $ is an $N \times N$ diagonal matrix with $(i,i)^{th}$ entry corresponding to $X_t=i$ and equal to $f^{(i)}\left(y_t;\theta^{(i)}\right)\calL_{\text{fine}}\left(\theta^{*(i)};y^*_t\right)$. 
%
Possible candidates for the fine-scale model include any of the models described in the previous subsections (HMM, CarHMM, HHMM, and HHMM-DFT) in addition to many others not described here.

For example, \citet{Bebbington:2007} and \citet{Borchers:2013} investigate data sets with count onsets as observations, so they use variations on a Poisson process as their fine-scale model. If the fine-scale model is a simple Poisson process, then the hierarchical structure described here reduces to a Markov-modulated Poisson process \citep{Fischer:1993}.
Further, \citep{Langrock:2018} use B-splines to model the emission distribution of an HMM as well as incorporate covariates. This non-parametric approach admits a penalized likelihood which can easily be incorporated into the hierarchical structure here. 
One additional class of fine-scale models is the set of continuous time methods such as the continuous-time HMM (CTHMM) \citep{Liu:2015} or the state-switching Ornstein-Uhlenbeck (OU) process \citep{Michelot:2019}. 
A continuous-time HMM (CTHMM) may be appropriate if observations are not equi-spaced in time \citep{Liu:2015}. As an example, \citet{Xu:2018} explain biologging accelerometer data by incorporating a CTHMM into a hierarchical model similar to ours. However, they take a longitudinal approach whereas we use a coarse-scale HMM in our case study. We also show the appendix that the state-switching OU process of \citet{Michelot:2019} is equivalent to a CarHMM under certain conditions.

These examples are a few of many fine-scale models that can act as initial building blocks in a practitioner's toolbox to construct increasingly complex hierarchical models. As an example, consider a CarHHMM, in which the coarse-scale model is a traditional HMM and the fine-scale model is a CarHMM. Alternatively, one may construct an HHHMM, where three scales of observations are recorded ($Y$, $Y^*$, and $Y^{**}$) and traditional HMMs are used to model all three levels. It is clear that are a myriad of possible models can be built using this framework. However, these models quickly become complicated and should be constructed with care to achieve an adequate fit of the data while avoiding over-fitting and high computational costs.
